Comprehensive assessment of the climate-induced water scarcity over continental Chile

### 2023-05-30 article/article.Rmd
mejorar introduccion, estructura y empezar abstract

---
title: Comprehensive assessment of the climate-induced water scarcity over continental Chile
author:
  - name: Francisco Zambrano
    email: francisco.zambrano@umayor.com
    affiliation: Hemere Centro de Observación de la Tierra,Universidad Mayor
    correspondingauthor: true
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Some Institute of Technology
    organization: Big Wig University
    addressline: 1 main street
    city: Gotham
    state: State
    postcode: 123456
    country: United States
  - code: Another University
    organization: Department
    addressline: A street 29
    postcode: 2054 NX
    city: Manchester,
    country: The Netherlands
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  This is the abstract.

  It consists of two paragraphs.
keywords: 
  - keyword1
  - keyword2
journal: "An awesome journal"
date: "`r Sys.Date()`"
classoption: preprint, 3p
bibliography: references.bib
biblio-style: "apalike"
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

@@

In 2021, the sixth assessment report (AR6) from the working group I of the IPCC was released [@IPCC2021]. Chapter 11 [@IPCCCH112021] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes. The evidence has been strengthened since AR5 [@IPCC2013]. There is high confidence that the increasing global warning can expand the land area affected by increasing drought frequency and severity [@IPCCCH112021]. Chile has been facing a persistent rainfall deficit lasting for more than ten years [@Garreaud2017] which has impacted the hydrological system [@Boisier2018], and consequently the vegetation development [@Zambrano2020].

@@

Vegetation drought indices are proxies of productivity [@Paruelo2016;@Schucknecht2017]. The main environmental variables that affect it are water supply and demand [@Mishra2015].  We measure them by precipitation and evapotranspiration (ET), commonly collected from weather stations. Usually, in developing countries (i.e., Chile), incomplete records or gaps present a challenge. But, there are satellite estimates of these variables. To evaluate drought, the World Meteorological Organization (WMO; [@WMO2012]) has proposed the Standardized Precipitation Index (SPI; [@McKee1993]), a multiscalar drought index, which has been used worldwide. For Chile, @Zambrano2017 derived and evaluated it from the product of the Climate Hazards Group InfraRed Precipitation with Station data (CHIRPS; [@Funk2015]). For water demand, it is used ET. The vegetation biomass productivity is strongly related to ET [@FAO66]. The atmospheric evaporative demand (AED) represents the maximum ET rate from a land surface (without water restriction), also known as reference ET. The recommended method for its calculation is the FAO Penman-Monteith [@Pereira2015;@Allen2005]. Due to climate change, AED is increasing, driving ET rise [@IPCCCH112021]. But, it is not always true [@Milly2016]. For example, regions where AET is highest have the lowest ET. The MOD16 product [@Running2021;@Mu2011] provides AET and ET satellite estimates and has been used to derive drought indices [@Mu2013]. Soil moisture (SM) is an Essential Climate Variable (ECV) that modulates vegetative growth. The climate change initiative (CCI) from the European Space Agency (ESA) delivers the ESA CCI SM product [@Dorigo2017] (current version 6.1), which has been helpful to monitor drought [@Zhang2019]. Besides, total water storage can be retrieved by the Gravity Recovery and Climate Experiment (GRACE), which allows analyzing water availability changes [@Ahmed2014;@Ma2017]. The water demand and supply by remote sensing can help evaluate how they have impacted vegetation productivity.

The study aims to analyze the drought impact on vegetation through Central Chile for 2000-2020, using satellite data as proxies of productivity and water demand and supply. We will evaluate LULC change and use the persistent classes within 2001-2019 (> 80%) to analyze the zcNDVI index and its interconnection with precipitation deficit, AED, ET, and vegetation cover. Finally, we will investigate if the observed changes are linked to the TWS and SM.

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{../output/figs/map_study_con_landcover.png}
\caption{ (\textbf{a}) Location of Central Chile and zones north (NCCH), central (CCCH), and south (SCCH) Central Chile. (\textbf{b}) Topography reference map. (\textbf{c}) Land cover classes for 2019. (\textbf{d}) Persistent land cover classess (> 80\%) for 2001-2019.}
\end{figure}

@@

## Satellite data

## zcNDVI and cluster zones

## Landcover change and persistence

## Vegetation Cover

## Evapotranspiration, Soil moisture and water storage

## Data analysis

@@

## zcNDVI and cluster zones
<!-- \begin{figure}[H] -->
<!-- \includegraphics[width=\textwidth]{../figs/map_zcNDVI_ts_clusters.png} -->
<!-- \caption{ } -->
<!-- \end{figure} -->

## Landcover Change and persistent

```{r,echo=FALSE,fig.width =10,fig.pos='H'}
dataLCV_ts <- readRDS('../data/processed_data/timeseries_landcover_zone_LCclass_2001-2021.rds')  
dataLCV_trend <- readRDS('../data/processed_data/trends_landcover_2001-2021.rds')

@@

<!-- \begin{figure}[H] -->
<!-- \centering -->
<!-- \includegraphics[width=\textwidth]{../output/figs/LC_perZoneAndCluster.png} -->
<!-- \caption{ } -->
<!-- \end{figure} -->

## SPI, AED, and ET

## SM and Total Storage

# Discussion

>>>>

---
title: Comprehensive assessment of the climate-induced water scarcity over continental Chile
author:
  - name: Francisco Zambrano
    email: francisco.zambrano@umayor.com
    affiliation: Hemera Centro de Observación de la Tierra,Universidad Mayor
    correspondingauthor: true
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Hemera Centro de Observación de la Tierra
    organization: Facultad de Ciencias, Universidad Mayor.
    addressline: La Piramide 5750
    city: Santiago
    state: Huechuraba
    postcode: 123456
    country: United States
  - code: Another University
    organization: Department
    addressline: A street 29
    postcode: 2054 NX
    city: Manchester,
    country: The Netherlands
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  Human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes globally. Chile has been affected by persistent water scarcity which has been impacting the hydrological system and vegetation development. Central Chile it is been the focus of research studies due to the diminishing of water supply, nevertheless water deficit is expanded beyond, specifically to the south of the country.

keywords: 
  - keyword1
  - keyword2
journal: "Remote Sensing of Environment"
date: "`r Sys.Date()`"
classoption: preprint, 3p
bibliography: references.bib
biblio-style: "apalike"
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

@@

The sixth assessment report (AR6) from the working group I of the IPCC [@IPCC2021] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes and the evidence has been strengthened since AR5 [@IPCC2013]. There is high confidence that the increasing global warning can expand the land area affected by increasing drought frequency and severity [@IPCCCH112021]. Chile has been facing a persistent rainfall deficit lasting for more than ten years [@Garreaud2017] which has impacted the hydrological system [@Boisier2018], and consequently the vegetation development [@Zambrano2020].

@@

Vegetation drought indices are commonly used as proxies of productivity [@Paruelo2016;@Schucknecht2017]. The main environmental variables that affect productivity are water supply and demand [@Mishra2015].  Those are measured by precipitation and evapotranspiration (ET), commonly collected from weather stations. Usually, in developing countries (i.e., Chile), incomplete records or gaps present a challenge. But, there are satellite estimates of these variables. To evaluate drought, the World Meteorological Organization (WMO; [@WMO2012]) has proposed the Standardized Precipitation Index (SPI; [@McKee1993]), a multiscalar drought index, which has been used worldwide. For Chile, @Zambrano2017 derived and evaluated it from the product of the Climate Hazards Group InfraRed Precipitation with Station data (CHIRPS; [@Funk2015]). For water demand, it is used ET. The vegetation biomass productivity is strongly related to ET [@FAO66]. The atmospheric evaporative demand (AED) represents the maximum ET rate from a land surface (without water restriction), also known as reference ET. The recommended method for its calculation is the FAO Penman-Monteith [@Pereira2015;@Allen2005]. Due to climate change, AED is increasing, driving ET rise [@IPCCCH112021]. But, it is not always true [@Milly2016]. For example, regions where AET is highest have the lowest ET. The MOD16 product [@Running2021;@Mu2011] provides AET and ET satellite estimates and has been used to derive drought indices [@Mu2013]. Soil moisture (SM) is an Essential Climate Variable (ECV) that modulates vegetative growth. The climate change initiative (CCI) from the European Space Agency (ESA) delivers the ESA CCI SM product [@Dorigo2017] (current version 6.1), which has been helpful to monitor drought [@Zhang2019]. Besides, total water storage can be retrieved by the Gravity Recovery and Climate Experiment (GRACE), which allows analyzing water availability changes [@Ahmed2014;@Ma2017]. The water demand and supply by remote sensing can help evaluate how they have impacted vegetation productivity.

We aims to analyze the climate-induced water scarcity over continental Chile for 2000-2023 by using estimated environmental variables of biomass productivity, and water demand/supply gather from earth observation products. The specific objective for the study are i) to evaluate LULC change for 2001-2021, ii) to derive and  assess the vegetation drought index zcNDVI as a proxy of biomass productivity, iii) analyze the interconnection of zcNDVI with drought indices of supply (i.e., precipitation), demand (i.e, ET), and vegetation cover; and iv) we will investigate if the observed changes are linked to the TWS and SM.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{../output/figs/map_study_con_landcover.png}
\caption{ (\textbf{a}) Location of Central Chile and zones north (NCCH), central (CCCH), and south (SCCH) Central Chile. (\textbf{b}) Topography reference map. (\textbf{c}) Land cover classes for 2019. (\textbf{d}) Persistent land cover classess (> 80\%) for 2001-2021.}
\end{figure}

@@

## Data

### Satellite data

### in-situ data

## LULC change for 2001-2021

### Landcover change and persistence

## Drought index for biomass productivity zcNDVI

## Interconnection of productivity with drought indices of supply/demand and vegetation cover

## Totatl water storage and Soil Moisture

@@

## LULC change for 2001-2021

```{r,echo=FALSE,fig.width =10,fig.pos='H',warning=FALSE}

dataLCV_ts <- readRDS('../data/processed_data/timeseries_landcover_zone_LCclass_2001-2021.rds')  
dataLCV_trend <- readRDS('../data/processed_data/trends_landcover_2001-2021.rds')

@@

### Landcover change and persistence

## Drought index for biomass productivity zcNDVI

## Interconnection of productivity with drought indices of supply/demand and vegetation cover

## Total water storage and Soil Moisture

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{../output/figs/JPL_TELLUS_GRACE-FO_L3_Monthly_Land_WET_SMA_R61v04_macrozonas.png}
\caption{Total water storage (mm) trend for 2002-2023 in the five macrozones on continental Chile}
\end{figure}
# Discussion

### 2023-05-30 article/article.tex
mejorar introduccion, estructura y empezar abstract

  \journal{An awesome journal} % Sets Journal name

@@

  \title{Comprehensive assessment of the climate-induced water scarcity
over continental Chile}
    \author[Hemere Centro de Observación de la Tierra,Universidad
Mayor]{Francisco Zambrano%
  \corref{cor1}%
  \fnref{1}}
   \ead{francisco.zambrano@umayor.com} 
    \author[Another University]{Bob Security%
  %
  }
   \ead{bob@example.com} 
    \author[Another University]{Cat Memes%
  %
  \fnref{2}}
   \ead{cat@example.com} 
    \author[Some Institute of Technology]{Derek Zoolander%
  %
  \fnref{2}}
   \ead{derek@example.com} 
      \affiliation[Some Institute of Technology]{
    organization={Big Wig University},addressline={1 main
street},city={Gotham},postcode={123456},state={State},country={United
States},}
    \affiliation[Another University]{
    organization={Department},addressline={A street
29},city={Manchester,},postcode={2054 NX},country={The Netherlands},}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{This is the first author footnote.}
    \fntext[2]{Another author footnote.}

  \begin{abstract}
  This is the abstract.

  It consists of two paragraphs.
  \end{abstract}
    \begin{keyword}
    keyword1 \sep 
    keyword2
  \end{keyword}

@@

In 2021, the sixth assessment report (AR6) from the working group I of
the IPCC was released \citep{IPCC2021}. Chapter 11 \citep{IPCCCH112021}
indicates that human-induced greenhouse gas emissions have increased the
frequency and/or intensity of some weather and climate extremes. The
evidence has been strengthened since AR5 \citep{IPCC2013}. There is high
confidence that the increasing global warning can expand the land area
affected by increasing drought frequency and severity
\citep{IPCCCH112021}. Chile has been facing a persistent rainfall
deficit lasting for more than ten years \citep{Garreaud2017} which has
impacted the hydrological system \citep{Boisier2018}, and consequently
the vegetation development \citep{Zambrano2020}.

@@

Vegetation drought indices are proxies of productivity
\citep{Paruelo2016, Schucknecht2017}. The main environmental variables
that affect it are water supply and demand \citep{Mishra2015}. We
measure them by precipitation and evapotranspiration (ET), commonly
collected from weather stations. Usually, in developing countries (i.e.,
Chile), incomplete records or gaps present a challenge. But, there are
satellite estimates of these variables. To evaluate drought, the World
Meteorological Organization (WMO; \citep{WMO2012}) has proposed the
Standardized Precipitation Index (SPI; \citep{McKee1993}), a multiscalar
drought index, which has been used worldwide. For Chile,
\citet{Zambrano2017} derived and evaluated it from the product of the
Climate Hazards Group InfraRed Precipitation with Station data (CHIRPS;
\citep{Funk2015}). For water demand, it is used ET. The vegetation
biomass productivity is strongly related to ET \citep{FAO66}. The
atmospheric evaporative demand (AED) represents the maximum ET rate from
a land surface (without water restriction), also known as reference ET.
The recommended method for its calculation is the FAO Penman-Monteith
\citep{Pereira2015, Allen2005}. Due to climate change, AED is
increasing, driving ET rise \citep{IPCCCH112021}. But, it is not always
true \citep{Milly2016}. For example, regions where AET is highest have
the lowest ET. The MOD16 product \citep{Running2021, Mu2011} provides
AET and ET satellite estimates and has been used to derive drought
indices \citep{Mu2013}. Soil moisture (SM) is an Essential Climate
Variable (ECV) that modulates vegetative growth. The climate change
initiative (CCI) from the European Space Agency (ESA) delivers the ESA
CCI SM product \citep{Dorigo2017} (current version 6.1), which has been
helpful to monitor drought \citep{Zhang2019}. Besides, total water
storage can be retrieved by the Gravity Recovery and Climate Experiment
(GRACE), which allows analyzing water availability changes
\citep{Ahmed2014, Ma2017}. The water demand and supply by remote sensing
can help evaluate how they have impacted vegetation productivity.

The study aims to analyze the drought impact on vegetation through
Central Chile for 2000-2020, using satellite data as proxies of
productivity and water demand and supply. We will evaluate LULC change
and use the persistent classes within 2001-2019 (\textgreater{} 80\%) to
analyze the zcNDVI index and its interconnection with precipitation
deficit, AED, ET, and vegetation cover. Finally, we will investigate if
the observed changes are linked to the TWS and SM.

@@

\hypertarget{satellite-data}{%
\subsection{Satellite data}\label{satellite-data}}

\hypertarget{zcndvi-and-cluster-zones}{%
\subsection{zcNDVI and cluster zones}\label{zcndvi-and-cluster-zones}}

\hypertarget{landcover-change-and-persistence}{%
\subsection{Landcover change and
persistence}\label{landcover-change-and-persistence}}

\hypertarget{vegetation-cover}{%
\subsection{Vegetation Cover}\label{vegetation-cover}}

\hypertarget{evapotranspiration-soil-moisture-and-water-storage}{%
\subsection{Evapotranspiration, Soil moisture and water
storage}\label{evapotranspiration-soil-moisture-and-water-storage}}

\hypertarget{data-analysis}{%
\subsection{Data analysis}\label{data-analysis}}

@@

\hypertarget{zcndvi-and-cluster-zones-1}{%
\subsection{zcNDVI and cluster zones}\label{zcndvi-and-cluster-zones-1}}

\hypertarget{landcover-change-and-persistent}{%
\subsection{Landcover Change and
persistent}\label{landcover-change-and-persistent}}

@@

\caption{\label{tab:unnamed-chunk-1}Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{r>{}r>{}r>{}r>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Trend of change [$km^2 year^{-1}$]} \\
\cmidrule(l{3pt}r{3pt}){2-7}
zone & Shrubland & Savanna & Grassland & Barren land & Forest & Cropland\\
\midrule
norte grande & -2.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c19134a3d.pdf} & 0.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c11281a1c.pdf} & -31.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c76e40d4c.pdf} & 32.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396cc832d9.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c22c3df0c.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c1e23c330.pdf}\\
norte chico & 79.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c3e39b63c.pdf} & -66.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c3a55945c.pdf} & -100.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c5ea10cfa.pdf} & 104.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c3e253181.pdf} & 0.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c6e3f818e.pdf} & -13.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c381a2e04.pdf}\\
zona central & 130.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c752a96cf.pdf} & -128.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c400c5e63.pdf} & 89.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c511304bc.pdf} & 23.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c58fe26f6.pdf} & -66.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c16afaacc.pdf} & -24.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c66febecf.pdf}\\
zona sur & -14.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c1ee86964.pdf} & -316.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c457bef18.pdf} & -55.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c688683cf.pdf} & 2.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c7f0e5ef0.pdf} & 412.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c4f819a2c.pdf} & 30.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c1671afb3.pdf}\\
zona austral & -44.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c7a04eff2.pdf} & 163.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c292ba1fa.pdf} & 226.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c362a0ad3.pdf} & -80.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c32966b1b.pdf} & -9.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c211e7d1e.pdf} & -1.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396cb504520.pdf}\\
\bottomrule
\end{tabular}}
\end{table}

\hypertarget{spi-aed-and-et}{%
\subsection{SPI, AED, and ET}\label{spi-aed-and-et}}

\hypertarget{sm-and-total-storage}{%
\subsection{SM and Total Storage}\label{sm-and-total-storage}}

>>>>

  \journal{Remote Sensing of Environment} % Sets Journal name

@@

  \title{Comprehensive assessment of the climate-induced water scarcity
over continental Chile}
    \author[Hemera Centro de Observación de la Tierra,Universidad
Mayor]{Francisco Zambrano%
  \corref{cor1}%
  \fnref{1}}
   \ead{francisco.zambrano@umayor.com} 
    \author[Another University]{Bob Security%
  %
  }
   \ead{bob@example.com} 
    \author[Another University]{Cat Memes%
  %
  \fnref{2}}
   \ead{cat@example.com} 
    \author[Some Institute of Technology]{Derek Zoolander%
  %
  \fnref{2}}
   \ead{derek@example.com} 
      \affiliation[Hemera Centro de Observación de la Tierra]{
    organization={Facultad de Ciencias, Universidad
Mayor.},addressline={La Piramide
5750},city={Santiago},postcode={123456},state={Huechuraba},country={United
States},}
    \affiliation[Another University]{
    organization={Department},addressline={A street
29},city={Manchester,},postcode={2054 NX},country={The Netherlands},}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{This is the first author footnote.}
    \fntext[2]{Another author footnote.}

  \begin{abstract}
  Human-induced greenhouse gas emissions have increased the frequency
  and/or intensity of some weather and climate extremes globally. Chile
  has been affected by persistent water scarcity which has been
  impacting the hydrological system and vegetation development. Central
  Chile it is been the focus of research studies due to the diminishing
  of water supply, nevertheless water deficit is expanded beyond,
  specifically to the south of the country.
  \end{abstract}
    \begin{keyword}
    keyword1 \sep 
    keyword2
  \end{keyword}

@@

The sixth assessment report (AR6) from the working group I of the IPCC
\citep{IPCC2021} indicates that human-induced greenhouse gas emissions
have increased the frequency and/or intensity of some weather and
climate extremes and the evidence has been strengthened since AR5
\citep{IPCC2013}. There is high confidence that the increasing global
warning can expand the land area affected by increasing drought
frequency and severity \citep{IPCCCH112021}. Chile has been facing a
persistent rainfall deficit lasting for more than ten years
\citep{Garreaud2017} which has impacted the hydrological system
\citep{Boisier2018}, and consequently the vegetation development
\citep{Zambrano2020}.

@@

Vegetation drought indices are commonly used as proxies of productivity
\citep{Paruelo2016, Schucknecht2017}. The main environmental variables
that affect productivity are water supply and demand \citep{Mishra2015}.
Those are measured by precipitation and evapotranspiration (ET),
commonly collected from weather stations. Usually, in developing
countries (i.e., Chile), incomplete records or gaps present a challenge.
But, there are satellite estimates of these variables. To evaluate
drought, the World Meteorological Organization (WMO; \citep{WMO2012})
has proposed the Standardized Precipitation Index (SPI;
\citep{McKee1993}), a multiscalar drought index, which has been used
worldwide. For Chile, \citet{Zambrano2017} derived and evaluated it from
the product of the Climate Hazards Group InfraRed Precipitation with
Station data (CHIRPS; \citep{Funk2015}). For water demand, it is used
ET. The vegetation biomass productivity is strongly related to ET
\citep{FAO66}. The atmospheric evaporative demand (AED) represents the
maximum ET rate from a land surface (without water restriction), also
known as reference ET. The recommended method for its calculation is the
FAO Penman-Monteith \citep{Pereira2015, Allen2005}. Due to climate
change, AED is increasing, driving ET rise \citep{IPCCCH112021}. But, it
is not always true \citep{Milly2016}. For example, regions where AET is
highest have the lowest ET. The MOD16 product
\citep{Running2021, Mu2011} provides AET and ET satellite estimates and
has been used to derive drought indices \citep{Mu2013}. Soil moisture
(SM) is an Essential Climate Variable (ECV) that modulates vegetative
growth. The climate change initiative (CCI) from the European Space
Agency (ESA) delivers the ESA CCI SM product \citep{Dorigo2017} (current
version 6.1), which has been helpful to monitor drought
\citep{Zhang2019}. Besides, total water storage can be retrieved by the
Gravity Recovery and Climate Experiment (GRACE), which allows analyzing
water availability changes \citep{Ahmed2014, Ma2017}. The water demand
and supply by remote sensing can help evaluate how they have impacted
vegetation productivity.

We aims to analyze the climate-induced water scarcity over continental
Chile for 2000-2023 by using estimated environmental variables of
biomass productivity, and water demand/supply gather from earth
observation products. The specific objective for the study are i) to
evaluate LULC change for 2001-2021, ii) to derive and assess the
vegetation drought index zcNDVI as a proxy of biomass productivity, iii)
analyze the interconnection of zcNDVI with drought indices of supply
(i.e., precipitation), demand (i.e, ET), and vegetation cover; and iv)
we will investigate if the observed changes are linked to the TWS and
SM.

@@

\hypertarget{data}{%
\subsection{Data}\label{data}}

\hypertarget{satellite-data}{%
\subsubsection{Satellite data}\label{satellite-data}}

\hypertarget{in-situ-data}{%
\subsubsection{in-situ data}\label{in-situ-data}}

\hypertarget{lulc-change-for-2001-2021}{%
\subsection{LULC change for 2001-2021}\label{lulc-change-for-2001-2021}}

\hypertarget{landcover-change-and-persistence}{%
\subsubsection{Landcover change and
persistence}\label{landcover-change-and-persistence}}

\hypertarget{drought-index-for-biomass-productivity-zcndvi}{%
\subsection{Drought index for biomass productivity
zcNDVI}\label{drought-index-for-biomass-productivity-zcndvi}}

\hypertarget{interconnection-of-productivity-with-drought-indices-of-supplydemand-and-vegetation-cover}{%
\subsection{Interconnection of productivity with drought indices of
supply/demand and vegetation
cover}\label{interconnection-of-productivity-with-drought-indices-of-supplydemand-and-vegetation-cover}}

\hypertarget{totatl-water-storage-and-soil-moisture}{%
\subsection{Totatl water storage and Soil
Moisture}\label{totatl-water-storage-and-soil-moisture}}

@@

\hypertarget{lulc-change-for-2001-2021-1}{%
\subsection{LULC change for
2001-2021}\label{lulc-change-for-2001-2021-1}}

@@

\caption{\label{tab:unnamed-chunk-1}Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{r>{}r>{}r>{}r>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Trend of change [$km^2 year^{-1}$]} \\
\cmidrule(l{3pt}r{3pt}){2-7}
zone & Shrubland & Savanna & Grassland & Barren land & Forest & Cropland\\
\midrule
norte grande & -2.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671073045199.pdf} & 0.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671013721869.pdf} & -31.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671079e41f67.pdf} & 32.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671075ebd9d8.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710524640a1.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671012073e8a.pdf}\\
norte chico & 79.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671061580771.pdf} & -66.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967101832bef9.pdf} & -100.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967105771ba87.pdf} & 104.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671042ea518f.pdf} & 0.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710607baab3.pdf} & -13.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967106f91099.pdf}\\
zona central & 130.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967105c411aeb.pdf} & -128.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967108a6119a.pdf} & 89.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671016b408b2.pdf} & 23.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967103ee6f936.pdf} & -66.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671074c30c6.pdf} & -24.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671056667823.pdf}\\
zona sur & -14.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967106cc832ad.pdf} & -316.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671034e3692f.pdf} & -55.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967104ac90971.pdf} & 2.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967107f7df47e.pdf} & 412.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967104e385249.pdf} & 30.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671013698bfb.pdf}\\
zona austral & -44.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710491ed82a.pdf} & 163.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710190c87cf.pdf} & 226.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967101d6ab929.pdf} & -80.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671033cd1ca4.pdf} & -9.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967101f22c71d.pdf} & -1.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710336ead1.pdf}\\
\bottomrule
\end{tabular}}
\end{table}

\hypertarget{landcover-change-and-persistence-1}{%
\subsubsection{Landcover change and
persistence}\label{landcover-change-and-persistence-1}}

\hypertarget{drought-index-for-biomass-productivity-zcndvi-1}{%
\subsection{Drought index for biomass productivity
zcNDVI}\label{drought-index-for-biomass-productivity-zcndvi-1}}

\hypertarget{interconnection-of-productivity-with-drought-indices-of-supplydemand-and-vegetation-cover-1}{%
\subsection{Interconnection of productivity with drought indices of
supply/demand and vegetation
cover}\label{interconnection-of-productivity-with-drought-indices-of-supplydemand-and-vegetation-cover-1}}

\hypertarget{total-water-storage-and-soil-moisture}{%
\subsection{Total water storage and Soil
Moisture}\label{total-water-storage-and-soil-moisture}}

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{../output/figs/JPL_TELLUS_GRACE-FO_L3_Monthly_Land_WET_SMA_R61v04_macrozonas.png}
\caption{Total water storage (mm) trend for 2002-2023 in the five macrozones on continental Chile}
\end{figure}
